\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \footnote{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-02-02 move_schedules.tex
overleaf changes from jose

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title[Fast Persistence Computations in Dynamic Settings]{
	\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings
	\thanks{
	\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.
	}
}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
With state of the art tools, there is no recourse when faced with a time-varying complex containing millions of simplices across thousands of snapshots in time.
Moreover, acquiring such a capability has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes all immediately become computationally viable tools in dynamic settings. Persistence-based examples include adaptive template functions~\cite{polanco2019adaptive}, persistence images~\cite{adams2017persistence}, and $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time via a decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
In particular, the vineyards algorithm transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc. see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 does have an implementation of vineyards, however the algorithm was never ported to version 2. Other major packages, such as GUDHI and PHAT, do not have vineyards implementations.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, they observed maintaining the decomposition along a certain parameterized family is the most computationally demanding aspect of RIVET~\cite{rivet}, software for computing and visualizing	 two-parameter persistent homology.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling a \emph{continuous} 1-parameter family of diagrams, it is not necessarily efficient when the parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ Code is available here: \url{https://github.com/peekxc/dart}}

@@

Contrasting the dynamic setting, there is extensive work  on improving the efficiency of computing a single (static) $R = DV$ decomposition. Chen~\cite{chen2011persistent} proposed \emph{persistence with a twist}, also called the \emph{clearing optimization}, which exploits a boundary/cycle relationship to ``kill'' columns early in the reduction rather than reducing them. 
Another popular optimization is to utilize the duality between   homology and cohomology~\cite{de2011dualities}, which dramatically improves the effectiveness of the clearing optimization~\cite{bauer2021ripser}. 
There are many other optimizations on the implementation side: the use of ranking functions defined on the combinatorial number system enables implicit cofacet enumeration, removing the need to store the boundary matrix explicitly; the apparent/emergent pairs optimization identifies columns whose pivot entries are unaffected by the reduction algorithm, reducing the total number of columns which need be reduced; sparse data structures such as bit-trees and lazy heaps allow for efficient column-wise additions with $\mathbb{Z}_2 = \mathbb{Z}/2\mathbb{Z}$ coefficients and effective $O(1)$ pivot entry retrieval, and so on~\cite{bauer2021ripser, bauer2017phat}.

@@

If  zeroth homology ($p = 0$) is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the relations are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give an algorithm that maintains a \emph{merge tree} with $e$ edges in $O(e)$ time per-update.
If only Betti numbers are needed, the zeroth-dimension problem reduces even further to the \emph{dynamic connectivity problem}, which can be be efficiently solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

 \subsection{A Motivating Example}\label{sec:motivation} 
We illustrate why the vineyards algorithm does not always yield an efficient strategy for time-varying setting with a simple experiment. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid and its associated sublevel-set filtrations:   
\begin{center}
	\vspace{1em}
	\includegraphics[height=1.25in]{circle_complex.png}
\end{center}
Each image contains a fixed number of pixels whose intensities vary with time. For each image, we build a fixed-size filtered simplicial complex using the \emph{Freudenthal} triangulation of the plane whose inclusion maps are obtained by lower stars of pixel values.   
%Snapshots of the image data and its corresponding time-varying filtration are shown below Figure~\ref{fig:vineyards}. 
%\begin{figure*}[h!]
%\centering	
%\end{figure*}
%Note that the size of the filtration is fixed---no simplices are added or removed over time. 
Two events critically change the pairing function: the first occurs when the central connected component splits to form a cycle, and the second  when the annulus  splits into four components. 
From left to right, the Betti numbers of the five evenly spaced `snapshots' of the filtration shown above are: $(\beta_0,\beta_1)  = (1,0), \; (1,1),  \; (1,1), \; (1,1), \; (4,0)$.
Thus, in this example, only a few persistence diagrams are needed to capture the major changes to the topology. 
\begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.75\textwidth]{circle_vineyards_vector.pdf}
 	\caption{The cumulative column operations needed to compute persistence across the time-varying filtration of grayscale images. Observe 10 independent persistence computations evenly spaced in time (green line) captures the major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}

  We use this data set as a baseline for comparing vineyards and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}). Suppose a practitioner wanted to know the major homological changes a time-varying filtration encounters over time.
  Since it is unknown a priori at when the persistent pairing function changes, one solution is to do $n$ independent persistence computations at $n$ evenly spaced points in the time domain.  An alternative approach is to construct a homotopy between a pair of filtrations $(K, f)$, $(K,f')$ and then decompose this homotopy into adjacent transpositions of the filtration order---the vineyards approach. 
 We refer to the former as the \emph{discrete setting}, which is often used in practice, and the  latter as the \emph{continuous setting}. Note that though the discrete setting is often more practical, it is only not guaranteed to capture all homological changes in persistence that occur in simulating a continuous 1-parameter family of diagrams.

 The cumulative cost (in total column operations) of these various approaches are shown on the left side of Figure~\ref{fig:vineyards}, wherein the reduction algorithm (\texttt{pHcol}) and the vineyards algorithm are compared. Two discrete strategies (green and purple) and two continuous strategies (black and blue) are shown.
  Note that without knowing where persistence pairing function changes, a continuous strategy must construct all $\approx 7 \times 10^{4}$ diagrams induced by the homotopy. 
  In this setting, as shown in the figure, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than vineyards, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards framework must incur some overhead (in terms of column operations) to maintain the underlying decomposition, even when pairing function is unchanged. 
Indeed, as shown by the case where $n = 10$, applying \texttt{pHcol} independently between  relatively ``close'' filtrations is substantially more efficient than iteratively updating the decomposition.
 Moreover, any optimizations to the reduction algorithm (e.g. clearing~\cite{chen2011persistent}) would only increase this disparity.
Since persistence has found many applications in dynamic contexts~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}, a more efficient alternative to vineyards is clearly needed. 
 \\
 \\
 \noindent
\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et al.~\cite{busaryev2010tracking} to include  coarser  operations---in terms of the number of valid intermediate decomposition states, as compared to vineyards---for dynamic persistence. 
We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case sizes of these quantities in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistence computations is given, wherein move schedules are demonstrated to be an order of magnitude more efficient than existing approaches at calculating persistence in dynamic settings. In particular, we illustrate the effectiveness of efficient scheduling with a variety of real-world applications, including flock analysis in dynamic metric spaces and manifold detection from image data using 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

\subsection{Main results}
Given a simplicial complex $K$ with filtration function $f : K \to [m]$, denote by $R = D V$ the decomposition of its corresponding boundary matrix $D$ such that $R$ is reduced and $V$ is upper-triangular (see section~\ref{sec:reduction} for details).
 If one has a pair of filtrations $(K, f)$,  $(K, f')$ of size $m = \lvert K \rvert$ and $R = D V$ has been computed for $K$, then it may be advantageous to use the information stored in ($R$, $V$) to reduce the computation of $R' = D' V'$. 
Given a permutation $P$ such that $D' = P^T D P$, such an update scheme has the form: 
$$ (\ast P^T \ast R \ast P \ast) = (P^T DP)(\ast P^T \ast V \ast P \ast) $$
where $\ast$ is substituted with elementary column operations that repair the permuted decomposition. It is known how to linearly interpolate $f \mapsto f'$ using $d \sim O(m^2)$ updates to the decomposition, where each update  requires at most two column operations~\cite{cohen2006vines}. Since each column operation takes $O(m)$, the complexity of reindexing $f \mapsto f'$ is $O(m^3)$, which is efficient if all $d$ decompositions are needed. Otherwise, if only $(R', V')$ is needed (and none of the intermediate $d -1$ decompositions), updating $R \mapsto R'$ using the approach from~\cite{cohen2006vines} matches the complexity of computing $R' = D' V'$ independently.

We now summarize the main results obtained in this effort (Theorem~\ref{prop:sparsity_result}): suppose one has a sequence of permutations $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ yielding a corresponding sequence of decompositions:
  \begin{equation}\label{eq:rv_seq}
   	R = R_0 = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R_d = R'
 \end{equation}
 where each permutation $s_i$ is a particular type of cyclic permutation (see section~\ref{sec:moves_setting}). If $i_\ast < j_\ast$, our first result extends~\cite{busaryev2010tracking} by showing that~\eqref{eq:rv_seq} can be computed in using $O(\kappa)$ column operations, where: 
\begin{equation}
	\quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert 
\end{equation}
where the quantities $\lvert \mathbb{I}_{k} \rvert$ and $\lvert \mathbb{J}_{k} \rvert$ depend on the sparsity of the $V_k$ and $R_k$ matrices, respectively, and $d \sim O(m)$ is a constant that depends on how similar $f$ and $f'$ are. The advantage of this result is that it depends explicitly on the sparsity pattern of the decomposition itself and is thus output sensitive, which we take advantage of in Section~\ref{sec:schedule_cost}. 
%Surprisingly, due to a refined analysis of restore subprocedure from~\cite{busaryev2010tracking} we give in section~\ref{sec:continuous_setting}, we find that move operations are actually cheaper than vineyards, see Proposition~\ref{prop:factor2}.
%if the vineyards algorithm requires $T$ column operations to obtain~\eqref{eq:rv_seq}, we show (Proposition~\ref{prop:factor2}) that obtains the same sequence using at most $M$ column operations, where: 
%\begin{equation}
%	M \leq \tfrac{1}{2}  T
%\end{equation}

Our second result turns towards upper bounding $d = \lvert \mathcal{S} \rvert$ and the complexity of constructing $\mathcal{S}$ itself. 
%The complexity of determining $\mathcal{S}$ depends on the type of permutations considered; many edit distances exhibit tight quadratic growth rates and tend to require at least quadratic time (or worse) to compute~\cite{labarre2013lower}. 
By reinterpreting a special set of cyclic permutations as edit operations on strings, we find that the \emph{smallest} such sequence that maps $f \mapsto f'$ has size $d$ (Proposition~\ref{prop:schedule_size}), where: 
\begin{equation}
	d = m - \lvert \mathrm{LCS}(f, f') \rvert 
\end{equation}
We also show that the information needed to construct any $\mathcal{S}$ with optimal size can be computed in $O(m \log \log m)$ preprocessing time and $O(m)$ memory. Although $d$ clearly is $O(m)$ for pathological inputs, we give evidence  that $d \sim m - \sqrt{m}$ in expectation for random filtrations is not too pessimistic for truly random filtrations (Corollary~\ref{cor:expectation}), and we give empirical results suggesting $d$ is tends to be much smaller for time-varying filtrations.

%Finally, we perform extensive experiments 
% given a decomposition $R = \partial_{K} V$ of a simplicial filtration $K$, algorithm~\ref{alg:schedule} can produce such a sequence $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$
%

@@

\textbf{Example 2.1: Reduction} Consider a triangle with vertices $u,v,w$, edges $a = (u,w)$, $b = (v,w)$, $c = (u, v)$,

@@

Using $\mathbb{Z}_2$ coefficients for simplicity, the reduction proceeds to compute $H_1(K)$ as follows:

>>>>

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title[Fast Persistence Computations in Dynamic Settings]{
	\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings
	\footnote{
	\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.
	}
}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
With state of the art tools, there is no recourse when faced with a time-varying complex containing millions of simplices across thousands of snapshots in time.
Moreover, acquiring such a capability has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes all immediately become computationally viable tools in dynamic settings. Such persistence summaries include adaptive template functions~\cite{polanco2019adaptive}, persistence images~\cite{adams2017persistence}, and $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a filtered simplicial complex $K$ with $m$ simplices in $O(m^3)$ time, via a particular decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm, in turn,  transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc. see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 does have an implementation of vineyards, however the algorithm was never ported to version 2. Other major packages, such as GUDHI and PHAT, do not have vineyards implementations.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, they observe that maintaining the decomposition along a certain parameterized family is the most computationally demanding aspect of RIVET~\cite{rivet}, a software for computing and visualizing	 two-parameter persistent homology.

The work presented here seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling a \emph{continuous} 1-parameter family of diagrams, it is not necessarily efficient when the parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ Code is available here: \url{https://github.com/peekxc/dart}}

@@

Contrasting the dynamic setting, there is extensive work  on improving the efficiency of computing a single (static) $R = DV$ decomposition. Chen~\cite{chen2011persistent} proposed \emph{persistence with a twist}, also called the \emph{clearing optimization}, which exploits a boundary/cycle relationship to ``kill'' columns early in the reduction rather than reducing them. 
Another popular optimization is to utilize the duality between   homology and cohomology~\cite{de2011dualities}, which dramatically improves the effectiveness of the clearing optimization~\cite{bauer2021ripser}. 
There are many other optimizations on the implementation side: the use of ranking functions defined on the combinatorial number system enables implicit cofacet enumeration, removing the need to store the boundary matrix explicitly; the apparent/emergent pairs optimization identifies columns whose pivot entries are unaffected by the reduction algorithm, reducing the total number of columns which need to be reduced; sparse data structures such as bit-trees and lazy heaps allow for efficient column-wise additions with $\mathbb{Z}_2 = \mathbb{Z}/2\mathbb{Z}$ coefficients and effective $O(1)$ pivot entry retrieval, and so on~\cite{bauer2021ripser, bauer2017phat}.

@@

If  zeroth homology   is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the relations are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give an algorithm that maintains a \emph{merge tree} with $e$ edges in $O(e)$ time per-update.
If only Betti numbers are needed, the zeroth-dimension problem reduces even further to the \emph{dynamic connectivity problem}, which can be be efficiently solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

 \subsection{A Motivating Example}\label{sec:motivation} 
We  will use a simple experiment to illustrate why the vineyards algorithm does not always yield an efficient strategy for time-varying settings. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, and its associated sublevel-set filtrations, as shown in Figure~\ref{fig:vidExample}.

\begin{figure}[!htb]
    \centering
    \includegraphics[height=1.25in]{circle_complex.png}
    \caption{Top: A video of an expanding annulus. Bottom: Sublevel-set filtrations, via pixel intensity, of a Freudenthal triangulation of the plane.}
    \label{fig:vidExample}
\end{figure}

Each image contains a fixed number of pixels whose intensities vary with time. For each image, we build a fixed-size filtered simplicial complex using the \emph{Freudenthal} triangulation of the plane whose inclusion maps are obtained by lower stars of pixel values.   
%Snapshots of the image data and its corresponding time-varying filtration are shown below Figure~\ref{fig:vineyards}. 
%\begin{figure*}[h!]
%\centering	
%\end{figure*}
%Note that the size of the filtration is fixed---no simplices are added or removed over time. 
Two events critically change the persistence diagrams: the first occurs when the central connected component splits to form a cycle, and the second  when the annulus  splits into four components. 
From left to right, the Betti numbers of the five evenly spaced `snapshots' of the filtration shown above are: $(\beta_0,\beta_1)  = (1,0), \; (1,1),  \; (1,1), \; (1,1), \; (4,0)$.
Thus, in this example, only a few persistence diagrams are needed to capture the major changes to the topology.

We use this data set as a baseline for comparing vineyards and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}). Suppose a practitioner wanted to know the major homological changes a time-varying filtration encounters over time.
Since it is unknown a priori when the persistent pairing function changes, one solution is to do $n$ independent persistence computations at $n$ evenly spaced points in the time domain.  An alternative approach is to construct a homotopy between a pair of filtrations $(K, f)$, $(K,f')$ and then decompose this homotopy into adjacent transpositions based on the filtration order---the vineyards approach. 
 We refer to the former as the \emph{discrete setting}, which is often used in practice, and the  latter as the \emph{continuous setting}. Note that though the discrete setting is often more practical, it is not guaranteed to capture all homological changes in persistence that occur in simulating a continuous 1-parameter family of diagrams.

The cumulative cost (in total column operations) of these various approaches are shown in Figure~\ref{fig:vineyards}, wherein the reduction (\texttt{pHcol}) and vineyard algorithms are compared. Two discrete strategies (green and purple) and two continuous strategies (black and blue) are shown.

\begin{figure*}[!htb]
 	\centering
 	\includegraphics[width=0.75\textwidth]{circle_vineyards_vector.pdf}
 	\caption{The cumulative column operations needed to compute persistence across the time-varying filtration of grayscale images. Observe 10 independent persistence computations evenly spaced in time (green line) captures the major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}

Note that without knowing where persistence pairing function changes, a continuous strategy must construct all $\approx 7 \times 10^{4}$ diagrams induced by the homotopy. 
In this setting, as shown in the figure, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than vineyards, while still capturing the main events.

The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards framework must incur some overhead (in terms of column operations) to maintain the underlying decomposition, even when the pairing function determining the persistence diagram is unchanged. 
Indeed, as shown by the case where $n = 10$, applying \texttt{pHcol} independently between  relatively ``close'' filtrations is substantially more efficient than iteratively updating the decomposition.
 Moreover, any optimizations to the reduction algorithm (e.g. clearing~\cite{chen2011persistent}) would only increase this disparity.
Since persistence has found many applications in dynamic contexts~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}, a more efficient alternative to vineyards is clearly needed. 
 \\
 \\
 \noindent
\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et al.~\cite{busaryev2010tracking} to include  coarser  operations for dynamic persistence settings. 
By a reduction to an edit distance problem, we give a   lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition,  along with a proof of its optimality.
We also give worst-case sizes of these quantities in expectation as well as efficient algorithms for constructing these operations---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
These operations parameterize sequences of cyclic permutations $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for heuristically solving this proxy minimization. 
A performance comparison with other reduction-based persistence computations is given, wherein move schedules are demonstrated to be an order of magnitude more efficient than existing approaches at calculating persistence in dynamic settings. In particular, we illustrate the effectiveness of efficient scheduling with a variety of real-world applications, including flock analysis in dynamic metric spaces and manifold detection from image data using 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

\subsection{Main results}
Given a simplicial complex $K$ with filtration function $f$, denote by $R = D V$ the decomposition of its corresponding boundary matrix $D$ such that $R$ is reduced and $V$ is upper-triangular (see section~\ref{sec:reduction} for details).
 If one has a pair of filtrations $(K, f)$,  $(K, f')$ of size $m = \lvert K \rvert$ and $R = D V$ has been computed for $(K,f)$, then it may be advantageous to use the information stored in ($R$, $V$) to reduce the computation of $R' = D' V'$. 
Given a permutation $P$ such that $D' = P D P^T$, such an update scheme has the form: 
$$ (\ast P \ast R \ast P^T \ast) = (P DP^T)(\ast P \ast V \ast P^T \ast) $$
where $\ast$ is substituted with elementary column operations that repair the permuted decomposition. It is known how to linearly interpolate $f \mapsto f'$ using $d \sim O(m^2)$ updates to the decomposition, where each update  requires at most two column operations~\cite{cohen2006vines}. Since each column operation takes $O(m)$, the complexity of reindexing $f \mapsto f'$ is $O(m^3)$, which is efficient if all $d$ decompositions are needed. Otherwise, if only $(R', V')$ is needed, updating $R \mapsto R'$ using the approach from~\cite{cohen2006vines} matches the complexity of computing $R' = D' V'$ independently.

We now summarize the main results obtained in this effort (Theorem~\ref{prop:sparsity_result}): suppose one has a sequence of permutations, i.e. a schedule,  $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ yielding a corresponding sequence of decompositions:
  \begin{equation}\label{eq:rv_seq}
   	R = R_0 = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R_d = R'
 \end{equation}
 where each permutation $s_k = (i_k, j_k)$ is a particular type of cyclic permutation (see section~\ref{sec:moves_setting}). If $i_\ast < j_\ast$ for all $s_\ast \in \mathcal{S}$, our first result extends~\cite{busaryev2010tracking} by showing that~\eqref{eq:rv_seq} can be computed  using $O(\kappa)$ column operations, where: 
\begin{equation}
	\quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert 
\end{equation}
The quantities $\lvert \mathbb{I}_{k} \rvert$ and $\lvert \mathbb{J}_{k} \rvert$ depend on the sparsity of the $V_k$ and $R_k$ matrices, respectively, and $d \sim O(m)$ is a constant that depends on how similar $f$ and $f'$ are. The advantage of this result is that it depends explicitly on the sparsity pattern of the decomposition itself and is thus output sensitive, which we take advantage of in Section~\ref{sec:schedule_cost}. 
%Surprisingly, due to a refined analysis of restore subprocedure from~\cite{busaryev2010tracking} we give in section~\ref{sec:continuous_setting}, we find that move operations are actually cheaper than vineyards, see Proposition~\ref{prop:factor2}.
%if the vineyards algorithm requires $T$ column operations to obtain~\eqref{eq:rv_seq}, we show (Proposition~\ref{prop:factor2}) that obtains the same sequence using at most $M$ column operations, where: 
%\begin{equation}
%	M \leq \tfrac{1}{2}  T
%\end{equation}

Our second result turns towards lower bounding $d = \lvert \mathcal{S} \rvert$ and the complexity of constructing $\mathcal{S}$ itself. 
%The complexity of determining $\mathcal{S}$ depends on the type of permutations considered; many edit distances exhibit tight quadratic growth rates and tend to require at least quadratic time (or worse) to compute~\cite{labarre2013lower}. 
By reinterpreting a special set of cyclic permutations as edit operations on strings, we find that the \emph{smallest} such sequence that maps $f \mapsto f'$ has size  (Proposition~\ref{prop:schedule_size}) : 
	d = m - \lvert \mathrm{LCS}(f, f') \rvert 
We also show that the information needed to construct any $\mathcal{S}$ with optimal size can be computed in $O(m \log \log m)$ preprocessing time and $O(m)$ memory. Although $d$ clearly is $O(m)$ for pathological inputs, we give evidence  that $d \sim m - \sqrt{m}$ in expectation for random filtrations is not too pessimistic for truly random filtrations (Corollary~\ref{cor:expectation}), and we give empirical results suggesting $d$ is tends to be much smaller for time-varying filtrations. 
%Finally, we perform extensive experiments 
% given a decomposition $R = \partial_{K} V$ of a simplicial filtration $K$, algorithm~\ref{alg:schedule} can produce such a sequence $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$

@@

\textbf{Example 2.1:} Consider a triangle with vertices $u,v,w$, edges $a = (u,w)$, $b = (v,w)$, $c = (u, v)$,

@@

Using $\mathbb{Z}_2$ coefficients, the reduction proceeds to compute $H_1(K)$ as follows:
